Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2025-01-10 paper/main.tex
revisions to GLMM methods section

In performing these analyses, our null hypothesis is that the knowledge
estimates we compute based on the quiz questions' embedding coordinates do
\textit{not} provide useful information about participants' abilities to answer
those questions---in other words, that there is no meaningful difference (on
average) between the knowledge estimates we compute for questions participants
answered correctly and those they answered incorrectly. Specifically, since we
estimate knowledge for a given embedding coordinate as a weighted
proportion-correct score (where each question's weight reflects its
embedding-space distance from the target coordinate; see Eqn.~\ref{eqn:prop}),
if these weights are uninformative (e.g., randomly distributed), then our
estimates of participants' knowledge should be equivalent (on average) to the
\textit{unweighted} proportion of correctly answered questions used to compute
them. In general, for a given participant and quiz, this expected value (i.e.,
that participant's proportion-correct score on that quiz) is the same for any
coordinate in the embedding space (e.g., any lecture timepoint, quiz question,
etc.). However, in the ``All questions'' and ``Within-lecture'' versions of the
analyses shown in Figure~\ref{fig:predictions}, we estimate each participant's
knowledge for each target question using all \textit{other} questions (or all
\textit{other} questions about the same lecture) they answered on the same quiz.
This introduces a systematic dependency between a participant's success on a
target question and their proportion-correct score on the questions available to
estimate their knowledge for it. For example, suppose a participant correctly
answered $n$ out of $q$ questions on a given quiz. If we hold out a single
\textit{correctly} answered question as the target, the proportion of remaining
questions answered correctly would be $\frac{n - 1}{q - 1}$. Whereas if we hold
out a single \textit{incorrectly} answered question, the proportion of remaining
questions answered correctly would be $\frac{n}{q - 1}$. Thus, the proportion of
correctly answered remaining questions (and therefore the null-hypothesized
value of a knowledge estimate computed from them) is always \textit{lower} for
target questions a participant answered correctly than for those they answered
incorrectly.

To correct for this baseline inverse relationship between a participant's
success on a target question and their estimated knowledge for it, we used a
rebalancing procedure that ensured our knowledge estimates for questions each
participant answered correctly and incorrectly were computed from the
\textit{same} proportion of correctly answered questions. For each target
question on a given participant's quiz, we identified all remaining questions
with the opposite ``correctness'' label (i.e., if the target question was
answered correctly, we identified all remaining incorrectly answered questions,
and vice versa). We then held out each of these opposite-label questions, in
turn, along with the target question, and estimated the participant's knowledge
for the target question using all \textit{other} remaining questions. Since each
of these subsets of remaining questions was constructed by holding out one
correctly answered question and one incorrectly answered question from the
participant's quiz, if the participant correctly answered $n$ out of $q$
questions total, then their proportion-correct score on each subset of questions
used to estimate their knowledge for the target question is $\frac{n-1}{q-2}$,
regardless of whether they answered the target question correctly or
incorrectly. Finally, averaging over these per-subset knowledge estimates
yielded a rebalanced estimate of the participant's knowledge for the target
question that leveraged information from all remaining questions' embedding
coordinates, but whose expected value under our null hypothesis was the same as
that of each individual subset ($\frac{n-1}{q-2}$). By equalizing the
null-hypothesized values of knowledge estimates for correctly and incorrectly
answered questions, this procedure ensures that any meaningful relationships we
observe between participants' estimated knowledge for individual quiz questions
and their ability to answer them are attributable to the predictive power of the
embedding-space distances used to weight questions' contributions to the
knowledge estimates, rather than an artifact of our estimation procedure. Note
that if a participant answered all or no questions on a given quiz correctly,
their responses contained no opposite-label questions with which to perform this
rebalancing, and we therefore excluded their data from our analyses for that
quiz. We used this rebalancing procedure when constructing knowledge estimates
for the ``All questions'' and ``Within-lecture'' versions of the analyses shown
in Figure~\ref{fig:predictions}, but not for the ``Across-lecture'' analyses as,
in this case, the target questions and the questions used to estimate
participants' knowledge for them were drawn from different subsets of quiz
questions (those about one lecture, and those about the other), and were
therefore independent.

In each version of this analysis (i.e., row in Fig.~\ref{fig:predictions}), and
separately for each of the three quizzes (i.e., column in
Fig.~\ref{fig:predictions}), we then fit a generalized linear mixed model
(GLMM) with a logistic link function to the set of knowledge estimates for all
questions that participants answered on the given quiz. We implemented these
models in \texttt{R} using the \texttt{lme4} package~\citep{BateEtal15a} and
fit them following guidance from~\citet{BateEtal15b} and \citet{MatuEtal17}.
Specifically, we initially fit each model with the maximal random effects
structure afforded by our design, which we identified as:
\[
    \mathtt{accuracy \sim knowledge + (knowledge\ \vert\ participant) + (knowledge\ \vert\ question)}
\]
where ``\texttt{accuracy}'' is a binary value indicating whether each target
question was answered correctly or incorrectly, ``\texttt{knowledge}'' is
estimated knowledge at each target question's embedding coordinate,
``\texttt{participant}'' is a unique identifier assigned to each participant,
and ``\texttt{question}'' is a unique identifier assigned to each quiz
question. For models we fit using knowledge estimates for target questions
about multiple content areas (i.e., in the ``All questions'' version of the
analysis), we also included an additional random effect term,
$\mathtt{(knowledge\ \vert\ lecture)}$, where ``\texttt{lecture}'' is a
categorical value denoting whether the target question was about \textit{Four
Fundamental Forces}, \textit{Birth of Stars}, or general physics knowledge.
Note that with our coding scheme, identifiers for each \texttt{question} are
implicitly nested within levels of \texttt{lecture} and do not require explicit
nesting in our model formula. We then iteratively removed random effects from
the maximal model until it successfully converged with a full rank (i.e., non-singular)
random effects variance-covariance matrix.

To assess the predictive value of our knowledge estimates, we compared each
GLMM's ability to discriminate between correctly and incorrectly answered
questions to that of an analogous model that did \textit{not} consider estimated
knowledge. Specifically, we used the same sets of observations with which we
fit each ``full'' model to fit a second ``null'' model, with the formula:
\[
    \mathtt{accuracy \sim (1\ \vert\ participant) + (1\ \vert\ question)}
\]
where ``\texttt{accuracy}'', ``\texttt{participant}'', and ``\texttt{question}'' are as defined above.
As with our full models, the null models we fit for the “All questions” version of the analysis for each quiz contained an additional term, $\mathtt{(1\ \vert\ lecture)}$, where ``\texttt{lecture}'' is as defined above.
We then compared each full model to its reduced (null) equivalent using a likelihood-ratio test (LRT).
Because the typical asymptotic $\chi^2_d$ approximation of the null distribution for the LRT statistic ($\lambda_{LR}$) is anti-conservative for models that differ in their random slope terms~\citep{GoldSimo00,ScheEtal08b,SnijBosk11}, we computed $p$-values for these tests using a parametric bootstrap procedure~\citep{DaviHink97,HaleHojs14}.
For each of 1,000 bootstraps, we used the fitted null model to simulate a sample of observations of equal size to our original sample.
We then re-fit both the null and full models to this simulated sample and compared them via an LRT.
This yielded a distribution of $\lambda_{LR}$ statistics we may expect to observe under our null hypothesis.
We computed a corrected $p$-value for our observed $\lambda_{LR}$ as $\frac{r+1}{n+1}$, where $r$ is the number of simulated model comparisons that yielded a $\lambda_{LR}$ greater than our observed value and $n$ is the number of simulations we ran (1,000).

>>>>

In performing these analyses, our null hypothesis is that the knowledge
estimates we compute based on the quiz questions' embedding coordinates do
\textit{not} provide useful information about participants' abilities to answer
those questions---in other words, that there is no meaningful difference (on
average) between the knowledge estimates we compute for questions participants
answered correctly and those they answered incorrectly. Specifically, since we
estimate knowledge for a given embedding coordinate as a weighted
proportion-correct score (where each question's weight reflects its
embedding-space distance from the target coordinate; see Eqn.~\ref{eqn:prop}),
if these weights are uninformative (e.g., randomly distributed), then our
estimates of participants' knowledge should be equivalent (on average) to the
\textit{unweighted} proportion of correctly answered questions used to compute
them. In general, for a given participant and quiz, this expected value (i.e.,
that participant's proportion-correct score on that quiz) is the same for any
coordinate in the embedding space (e.g., any lecture timepoint, quiz question,
etc.). However, in the ``All questions'' and ``Within-lecture'' versions of the
analyses shown in Figure~\ref{fig:predictions}, we estimate each participant's
knowledge for each target question using all \textit{other} questions (or all
\textit{other} questions about the same lecture) they answered on the same quiz.
This introduces a systematic dependency between a participant's success on a
target question and their proportion-correct score on the remaining questions
available to estimate their knowledge for it. For example, suppose a participant
correctly answered $n$ out of $q$ questions on a given quiz. If we hold out a
single \textit{correctly} answered question as the target, the proportion of
remaining questions answered correctly would be $\frac{n - 1}{q - 1}$. Whereas
if we hold out a single \textit{incorrectly} answered question, the proportion
of remaining questions answered correctly would be $\frac{n}{q - 1}$. Thus, the
proportion of correctly answered remaining questions (and therefore the
null-hypothesized value of a knowledge estimate computed from them) is always
\textit{lower} for target questions a participant answered correctly than for
those they answered incorrectly.

To correct for this baseline inverse relationship between a participant's
success on a target question and their estimated knowledge for it, we used a
rebalancing procedure that ensured our knowledge estimates for questions each
participant answered correctly and incorrectly were computed from the
\textit{same} proportion of correctly answered questions. For each target
question on a given participant's quiz, we identified all remaining questions
with the opposite ``correctness'' label (i.e., if the target question was
answered correctly, we identified all remaining incorrectly answered questions,
and vice versa). We then held out each of these opposite-label questions, in
turn, along with the target question, and estimated the participant's knowledge
for the target question using all \textit{other} remaining questions. Since each
of these subsets of remaining questions was constructed by holding out one
correctly answered question and one incorrectly answered question from the
participant's quiz, if the participant correctly answered $n$ out of $q$
questions total, then their proportion-correct score on each subset of questions
used to estimate their knowledge for the target question is $\frac{n-1}{q-2}$,
regardless of whether they answered the target question correctly or
incorrectly. Finally, averaging over these per-subset knowledge estimates
yielded a rebalanced estimate of the participant's knowledge for the target
question that leveraged information from all remaining questions' embedding
coordinates, but whose expected value under our null hypothesis was the same as
that of each individual subset ($\frac{n-1}{q-2}$). By equalizing the
null-hypothesized values of knowledge estimates for correctly and incorrectly
answered questions, this procedure ensures that any meaningful relationships we
observe between participants' estimated knowledge for individual quiz questions
and their abilities to correctly answer them are attributable to the predictive
power of the embedding-space distances used to weight questions' contributions
to the knowledge estimates, rather than an artifact of our estimation procedure.
Note that if a participant answered all or no questions on a given quiz
correctly, their responses contained no opposite-label questions with which to
perform this rebalancing, and we therefore excluded their data from our analyses
for that quiz. We used this rebalancing procedure when constructing knowledge
estimates for the ``All questions'' and ``Within-lecture'' versions of the
analyses shown in Figure~\ref{fig:predictions}, but not for the
``Across-lecture'' analyses as, in this case, the target questions and the
questions used to estimate participants' knowledge for them were drawn from
different subsets of quiz questions (those about one lecture, and those about
the other), and were therefore independent.

In each version of this analysis (i.e., row in Fig.~\ref{fig:predictions}), and
separately for each of the three quizzes (i.e., column in
Fig.~\ref{fig:predictions}), we then fit a generalized linear mixed model
(GLMM) with a logistic link function to the set of knowledge estimates for all
questions (or all questions about a particular lecture) that participants
answered on the given quiz. We implemented these models in \texttt{R} using the
\texttt{lme4} package~\citep{BateEtal15a} and fit them following guidance
from~\citet{BateEtal15b} and \citet{MatuEtal17}. Specifically, we initially fit
each model with the maximal random effects structure afforded by our design,
which we identified as:
\[
    \mathtt{accuracy \sim knowledge + (knowledge\ \vert\ participant) + (knowledge\ \vert\ question)}
\]
where ``\texttt{accuracy}'' is a binary value indicating whether each target
question was answered correctly or incorrectly, ``\texttt{knowledge}'' is
estimated knowledge at each target question's embedding coordinate,
``\texttt{participant}'' is a unique identifier assigned to each participant,
and ``\texttt{question}'' is a unique identifier assigned to each quiz
question. For models we fit using knowledge estimates for target questions
about multiple content areas (i.e., in the ``All questions'' version of the
analysis), we also included an additional random effect term,
$\mathtt{(knowledge\ \vert\ lecture)}$, where ``\texttt{lecture}'' is a
categorical value denoting whether the target question was about \textit{Four
Fundamental Forces}, \textit{Birth of Stars}, or general physics knowledge.
Note that with our coding scheme, identifiers for each \texttt{question} are
implicitly nested within levels of \texttt{lecture} and so do not require
explicit nesting in our model formula. We then iteratively removed random
effects from the maximal model until it successfully converged with a full-rank
random effects variance-covariance matrix. We obtained the odds ratios reported
in Figure~\ref{fig:predictions} by exponentiating the estimated coefficient for
``\texttt{knowledge}'' from each fitted model. Conceptually, these odds ratios
represent how many times greater the odds are that a given participant will
answer a given question correctly if their estimated knowledge for its embedding
coordinate is 1, compared to if it is 0. We estimated 95\% confidence intervals
for each odds ratio by generating 10,000 random subsamples (of full size, with
replacement) from the data used to fit each model, and refitting the models to
each subsample to obtain bootstrap distributions of 10,000 odds ratios.

To assess the predictive value of our knowledge estimates, we compared each
GLMM's ability to explain participants' success on individual quiz questions to
that of an analogous model which assumed (as we assume under our null hypothesis) that knowledge estimates for correctly and incorrectly answered
questions did \textit{not} systematically differ, on average. Specifically, we
used the same sets of observations with which we fit each ``full'' model to fit
a second ``null'' model that had the same random effects structure, but in which
the coefficient for the fixed effect of ''\texttt{knowledge}'' was fixed at 0
(i.e., we removed this term from the null model). We then compared each full
model to its reduced (null) equivalent using a likelihood-ratio test (LRT).
Because the standard asymptotic $\chi^2_d$ approximation of the null
distribution for the LRT statistic ($\lambda_{LR}$) can be anti-conservative for
finite sample sizes~\citep{GoldSimo00,ScheEtal08b,SnijBosk11}, we computed
$p$-values for these tests using a parametric bootstrap
procedure~\citep{DaviHink97,HaleHojs14}. For each of 10,000 bootstraps, we used
the fitted null model to simulate a sample of observations of equal size to our
original sample. We then re-fit both the null and full models to this simulated
sample and compared them via an LRT. This yielded a distribution of
$\lambda_{LR}$ statistics we may expect to observe given data that conforms to
our null hypothesis. We computed a corrected $p$-value for our observed
$\lambda_{LR}$ as $\frac{r+1}{n+1}$, where $r$ is the number of simulated model
comparisons that yielded a $\lambda_{LR}$ greater than our observed value and
$n$ is the number of simulations we ran (10,000).
